Revealing Transient Strain in Geodetic Data with the Radial Basis Function Finite Difference Method

### 2017-05-29 HinesHetlandGPR.tex
did more

% OUTLINE
% INTRODUCTION
% - Why we care about strain in the crust
%   - Knowing long term strain rates tells us where we can expect seismicity
%
% - How people have gone about estimating strain rates
%   - Draw distinction between parametric and nonparametric methods
%   - Draw distinction between model based and data based parametric methods 
%   - Talk about Kato
%
% - Another, more challenging problem is how to calculate time-dependent strain
%   - Time dependent strain rates could illuminate transient geophysical signals such as postseismic deformation, 
%	  slow slip events, or volcanic deformation. 	  
%	- Since numerous megathrust earthquakes were preceded by slow slip, it is important to detect these features.
%	- We can also use estimates of transient strain to calibrate strain meters.	
%	- People who have tackled this problem before include Ohtani and Holt.	
%	- A major deficiency with these methods is that they are parametric giving poor uncertainties. Without good 
%	  uncertainties we cannot objectively discern signal from noise and we cannot calibrate BSM.	
% - Here we propose GPR and we tackle several two major difficulties	
%   - The method is similar to an extension of Kato to the time domain. We address two debilitating difficulties that 
%     arise. We also dont do that autocovariance bullshit.  Instead we use REML.  
%	- outliers  
%	- dense matrix
%	
% - GPR can be used for more than just estimating strain and it is a powerful geophysical tool. We can use it to quantify
%   spatial noise in GPS data. (this should be a side blurb when talking about quantifying signal.   
%
% - We discuss the method, form prior with REML, and demonstrate utility in detecting SSE by comparing it to tremor.	  
%
% METHODS
% - define model
% - discuss known deficiencies with the model
% - describe covariance as a time-space separable function 
% - describe geophysically relevant gaussian processes. (include seasonal and spherical)
% - REML for determining optimal hyperparameters
% - conditioning the prior with observations
% - differentiating the Gaussian process
% - note on addition, subtraction, and scaling
% - REML
% - Numerical notes
%   - partition inverse and solve with CHOLMOD
%
% APPLICATION TO CASCADIA SSE
% - SSE background
% - REML results
%   - noise 
%   - signal for IBM, SE, and WEN
%   - Plot all covariances functions

@@

\subsection{Conditioning}
In this section we describe how to condition the prior for transient displacements, $u$, with GNSS observations, $\bm{d}_*$, to form a posterior estimate of transient displacements, $\hat{u} = u | \bm{d}_*$. We will assume that an appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_\epsilon$ have already been chosen. If $\lambda$ is kept finite, then we can follow \citet{Rasmussen2006} to find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(z) = C_u(z,\bm{y})\bm{C_d}^{-1}\bm{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(z,z') = C_u(z,\bm{y})\bm{C_d}^{-1}C_u(\bm{y},z').
\end{equation}
However, we are interested in the limit as $\lambda \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We can use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(z) =
\left[ 
\begin{array}{cc}
C_u(z,\bm{y}) & \bm{0} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\bm{y},\bm{y}) + C_\epsilon(\bm{y},\bm{y}) & \bm{P} \\
\bm{P}^T  & -\lambda^{-2} \bm{I} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
\bm{d}_* \\
\bm{0} \\
\end{array}
\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(z,z') = 
C_u(z,z') - 
\left[ 
\begin{array}{cc}
C_u(z,\bm{y}) & \bm{0} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\bm{y},\bm{y}) + C_\epsilon(\bm{y},\bm{y}) & \bm{P} \\
\bm{P}^T  & -\lambda^{-2} \bm{I} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
C_u(\bm{y},z') \\
\bm{0} \\
\end{array}
\right].
\end{equation}
Taking the limit as $\lambda \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(z) =
\left[ 
\begin{array}{cc}
C_u(z,\bm{y}) & \bm{0} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\bm{y},\bm{y}) + C_\epsilon(\bm{y},\bm{y}) & \bm{P} \\
\bm{P}^T  & \bm{0} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
\bm{d}_* \\
\bm{0} \\
\end{array}
\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(z,z') = 
C_u(z,z') - 
\left[ 
\begin{array}{cc}
C_u(z,\bm{y}) & \bm{0} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\bm{y},\bm{y}) + C_\epsilon(\bm{y},\bm{y}) & \bm{P} \\
\bm{P}^T  & \bm{0} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
C_u(\bm{y},z') \\
\bm{0} \\
\end{array}
\right].
\end{equation}
which allows us to evaluate the limit  evaluate the limit 
and we consider the GNSS observations $\bm{d}_*$ to be a realization of $\bm{d}$. Furthermore, we The random vector $\bm{d}$ is distributed as $\bm{d} \sim \mathcal{N}(\bar{\bm{d}},\bm{C_d})$, where 
\begin{equation}
\bm{d}
\end{equation}

>>>>

% OUTLINE
% INTRODUCTION
% - Why we care about strain in the crust
%   - Knowing long term strain rates tells us where we can expect seismicity
%
% - How people have gone about estimating strain rates
%   - Draw distinction between parametric and nonparametric methods
%   - Draw distinction between model based and data based parametric methods 
%   - Talk about Kato
%
% - Another, more challenging problem is how to calculate time-dependent strain
%   - Time dependent strain rates could illuminate transient geophysical signals such as postseismic deformation, 
%	  slow slip events, or volcanic deformation. 	  
%	- Since numerous megathrust earthquakes were preceded by slow slip, it is important to detect these features.
%	- We can also use estimates of transient strain to calibrate strain meters.	
%	- People who have tackled this problem before include Ohtani and Holt.	
%	- A major deficiency with these methods is that they are parametric giving poor uncertainties. Without good 
%	  uncertainties we cannot objectively discern signal from noise and we cannot calibrate BSM.	
% - Here we propose GPR and we tackle several two major difficulties	
%   - The method is similar to an extension of Kato to the time domain. We address two debilitating difficulties that 
%     arise. We also dont do that autocovariance bullshit.  Instead we use REML.  
%	- outliers  
%	- dense matrix
%	
% - GPR can be used for more than just estimating strain and it is a powerful geophysical tool. We can use it to quantify
%   spatial noise in GPS data. (this should be a side blurb when talking about quantifying signal.   
%
% - We discuss the method, form prior with REML, and demonstrate utility in detecting SSE by comparing it to tremor.	  
%
% METHODS
% - define model
% - discuss known deficiencies with the model
% - describe covariance as a time-space separable function 
% - describe geophysically relevant gaussian processes. (include seasonal and spherical)
% - REML for determining optimal hyperparameters
% - conditioning the prior with observations
% - differentiating the Gaussian process
% - note on addition, subtraction, and scaling
% - REML
% - Numerical notes
%   - partition inverse and solve with CHOLMOD
% - outlier detection

% APPLICATION TO CASCADIA SSE
% - SSE background
% - REML results
%   - noise 
%   - signal for IBM, SE, and WEN
%   - Plot all covariances functions

@@

\subsection{Conditioning}
In this section we describe how to condition the prior for transient displacements, $u$, with GNSS observations, $\bm{d}_*$, to form a posterior estimate of transient displacements, $\hat{u} = u | \bm{d}_*$. We will assume that an appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_\epsilon$ have already been chosen. If $\lambda$ is kept finite, then we can follow \citet{Rasmussen2006} to find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(z) = C_u(z,\bm{y})\bm{C_d}^{-1}\bm{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(z,z') = C_u(z,\bm{y})\bm{C_d}^{-1}C_u(\bm{y},z').
\end{equation}
However, we are interested in the limit as $\lambda \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We can use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(z) =
\left[ 
\begin{array}{cc}
C_u(z,\bm{y}) & \bm{0} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\bm{y},\bm{y}) + C_\epsilon(\bm{y},\bm{y}) & \bm{P} \\
\bm{P}^T  & -\lambda^{-2} \bm{I} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
\bm{d}_* \\
\bm{0} \\
\end{array}
\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(z,z') = 
C_u(z,z') - 
\left[ 
\begin{array}{cc}
C_u(z,\bm{y}) & \bm{0} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\bm{y},\bm{y}) + C_\epsilon(\bm{y},\bm{y}) & \bm{P} \\
\bm{P}^T  & -\lambda^{-2} \bm{I} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
C_u(\bm{y},z') \\
\bm{0} \\
\end{array}
\right].
\end{equation}
Taking the limit as $\lambda \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(z) =
\left[ 
\begin{array}{cc}
C_u(z,\bm{y}) & \bm{0} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\bm{y},\bm{y}) + C_\epsilon(\bm{y},\bm{y}) & \bm{P} \\
\bm{P}^T  & \bm{0} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
\bm{d}_* \\
\bm{0} \\
\end{array}
\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(z,z') = 
C_u(z,z') - 
\left[ 
\begin{array}{cc}
C_u(z,\bm{y}) & \bm{0} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\bm{y},\bm{y}) + C_\epsilon(\bm{y},\bm{y}) & \bm{P} \\
\bm{P}^T  & \bm{0} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
C_u(\bm{y},z') \\
\bm{0} \\
\end{array}
\right].
\end{equation}

The posterior transients displacements, $\hat{\bm{u}}$, are found by evaluating eq. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) for each displacement component. We differentiate the Gaussian process $\hat{\bm{u}}$ (see \cite{Abrahamsen1997}) to form an estimate of transient strain rate, $\dot{\bm{\epsilon}}$, which is itself a spatially and temporally continuous Gaussian process. The component of the transient strain rate tensor, $\dot{\epsilon}_{ij}$, are given by
\begin{equation}\label{eq:StrainRate}
\dot{\epsilon}_{ij}(z) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                     \frac{\partial \hat{u}_i(z)}{\partial x_j} +  
                                     \frac{\partial \hat{u}_j(z)}{\partial x_i}\right).
\end{equation}
Component $\dot{\epsilon}_{ij}$ of the strain rate tensor is distributed as $\mathcal{N}(\mu_{\dot{\epsilon}_{ij}},C_{\dot{\epsilon}_{ij}})$, where
\begin{equation}\label{eq:StrainMean}
\mu_{\dot{\epsilon}_{ij}}(z) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                               \frac{\partial \mu_{\hat{u}_i}(z)}{\partial x_j} + 
                               \frac{\partial \mu_{\hat{u}_j}(z)}{\partial x_i} \right)
\end{equation} 
and  
\begin{equation}\label{eq:StrainCov}
C_{\dot{\epsilon}_{ij}}(z,z') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                           \frac{\partial^2 C_{\hat{u}_i}(z,z')}{\partial x_j \, \partial x'_j} + 
                           \frac{\partial^2 C_{\hat{u}_j}(z,z')}{\partial x_i \, \partial x'_i} \right).
\end{equation}

which allows us to evaluate the limit  evaluate the limit 
and we consider the GNSS observations $\bm{d}_*$ to be a realization of $\bm{d}$. Furthermore, we The random vector $\bm{d}$ is distributed as $\bm{d} \sim \mathcal{N}(\bar{\bm{d}},\bm{C_d})$, where 
\begin{equation}
\bm{d}
\end{equation}
